Copulas and their potential for ecology

### 2019-10-01 Paper.Rmd
comments in prep for revisions for the Q2 sections#

Having demonstrated that non-normal copula structure and asymmetric tail dependence are 
common in ecological variables, we 
addressed Q2 by exploring, using models, three possible mechanisms
by which these phenomena may arise. 
The first mechanism relates to the ideas in the Introduction about 
Liebig's law of the minimum, and to nonlinear influences of environmental
variables on ecological variables.
If an environmental variables influences an ecological variable 
more strongly in one of its tails, we explored whether the ecological
variable could then exhibit asymmetric tail dependence across space.
Let $E_i(t)$ be an ecological variable 
measured at location $i$ ($i=1,2$) at time $t$.
Assume the dynamics 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where 
the $\delta_i(t)$ are standard-normally distributed 
and independent across time and locations, $a=0.2$,
$b=0.1, -0.1, 0.5, -0.5$ in different simulations, and 
the $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$.
The function $g$ describes how the environment $\epsilon_i(t)$ influences
$E_i(t)$. We used $g$ equal to $g_1$ or $g_2$ in different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise;
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that occur only below a threshold of $\epsilon=0$, 
and $g_2$ represents effects that occur only above a threshold of $\epsilon=0$.
Negative values of $b$ correspond to overcompensating dynamics 
and positive values to undercompensating dynamics; larger $|b|$
means slower return to equilibrium after a disturbance.

For each $b$ and $g_i$, we simulated the model for $25000$ time
steps and retained the $E_i(t)$ for the final $2500$ time 
steps. We applied our model selection algorithms and 
non-parametric statistics (see section on methods for Q1) to these 
outputs to discover if the model could
produce non-normal copula structure and asymmetric tail dependence in the 
$E_i(t)$. Because $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have normal copula structure, the Moran 
mechanism analyzed below 
does not operate here.

@@

To explore the importance of nonlinear dynamics for our proposed mechanism, 
we also considered a
nonlinear population model $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right)+\sigma\epsilon_i(t)\right)$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when it is strong, it brings the nonlinearities of the model into play.

For each of the model setups above, for each of several copula families, for each $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given Kendall
correlation $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were then computed for both noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted against $\tau$ for noises and populations. If the 
hypothesis was correct that characteristics of 
the copula structure of spatial dependence in an
ecological variable can be inherited from the copula structure of spatial 
dependence in an environmental variables, then 
plots should be similar for populations and noises.

The next mechanism we investigated is evolutionary, and pertains to bivariate 
trait data across species, e.g., our bird and mammal 
data. This mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
The hypothesis is that tail dependence occurs in evolutionary changes
in bivariate characters, and gives rise to tail dependence between the 
two character values across extant species.
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million 
independent draws from bivariate distributions showing one of five distinct types of copula 
structure: 1) extreme or 2) moderate left-tail dependence,
3) symmetric tail dependence, or 4) moderate or 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech}). 
All distributions had standard-normal marginals and Spearman 
correlation $0.875$ between components, so our simulations assess 
the impact of copula structure only. For 
each of the five copulas, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across phylogeny tips was assessed.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence in tip characters, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

Our model with asymmetric environmental effects 
produces outputs with visually apparent asymmetry 
of tail dependence, to an extent that depended on the value of $b$;
thus this is a mechanism that can produce non-normal copula structure and
asymmetric tail dependence across space in ecological variables.
This result is represented in Fig. \ref{MasterFigure} as the solid box around 
"Nonlinear environmental effects, Liebig's law" and the arrows labelled "A".
It is explained in the Discussion why the box and some of those arrows 
are solid, instead of dashed, although the results we 
present in this section are theoretical.
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ 
(Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetry of tail dependence was strong; for 
$b=\pm 0.5$, asymmetry was weaker but still apparent
(Figs \ref{SM-fig_ms_asym_b_0.5} - 
\ref{SM-fig_ms_asym_b_-0.5}).
Lower-tail (respectively, upper-tail) spatial dependence in 
effects of noise ($g_1$, respectively, $g_2$) 
produced lower-tail (respectively, upper-tail) spatial 
dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) confirmed that
ecological outputs had non-normal copula structure, but did not always reflect 
asymmetry of tail dependence, perhaps because those methods measured
dependence in the extreme tails. Results using our nonparametric statistics, however,
reflected the visually apparent asymmetry (Tables \ref{SM-tab_stat_npa_asym}).

@@

The hypothesis that characteristics of 
spatial copula structure and tail dependence
in an ecological variable can be inherited from an environmental 
variable was largely confirmed; 
however, similarities between environmental- and 
ecological-variable copula structure were reduced when 
dynamics were strongly nonlinear.
Thus this is a mechanism that can produce non-normal copula structure and
asymmetric tail dependence across space in ecological variables.
This result is represented in Fig. \ref{MasterFigure} as the dashed box around 
"Moran effects" and the arrows labelled "B".
For simulations using AR(1) models, copula 
statistics were always very similar for noise and populations 
(Fig \ref{fig_Cause4copula_scatter_C_kend_ar1}, 
\ref{SM-fig_Cause4copula_C_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1}),
though there were significant but minor differences for some statistics
and simulations.
For weak noise and our nonlinear model, copula statistics
were again very similar for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}).
Though there were again significant but small differences for some statistics
and simulations, the hypothesis was supported that copula 
structure in the noise translates
to similar but not identical copula structure in output populations. 
Since many ecological models are nonlinear, this result provides the 
expectation that there will be a Moran-effect-like correspondence 
between noise and model-output copula structure across space 
for typical ecological dynamics,
as long as noise is small enough that dynamics stay relatively close
to the model equilibrium. Results that hold for "weak noise" in this sense
are common in ecology. 
We repeated the analysis with $r=1.3$, with similar results. The deterministic 
one-habitat-patch Ricker model exhibits a monotonic approach to a stable equilibrium
at $K$ when $r<1$ (undercompensating dynamics, e.g., the value $r=0.5$ used initially), 
but exhibits an oscillatory approach to $K$ when $r>1$ (overcompensating
dynamics, e.g., $r=1.3$).

>>>>

Having demonstrated that non-normal copula structure and asymmetric tail associations are 
common in ecological variables, we 
addressed Q2 by exploring, using models, three possible mechanisms
by which these phenomena may arise. 
<!--***DAN: Maybe insert some text indicating these are first, exploratory models, and are not intended to comprehensively explore
a mechanism (that can be done in future work), but rather to get a sense
for whether observed copula structure may be partly due to the
mechanism, and whether the mechanism is worthy of more comprehensive
exploration in that connection in future work. One way these are noncomprehensive
explorations is that I chose particular limited sets of parameters.-->
The first mechanism relates to the ideas in the Introduction about 
Liebig's law of the minimum, and to nonlinear influences of environmental
variables on ecological variables.
If an environmental variables influences an ecological variable 
more strongly in one of its tails, we explored whether the ecological
variable could then exhibit asymmetric tail dependence across space.
Let $E_i(t)$ be an ecological variable 
measured at location $i$ ($i=1,2$) at time $t$.
Assume the dynamics 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where 
the $\delta_i(t)$ are standard-normally distributed 
and independent across time and locations, $a=0.2$,
$b=0.1, -0.1, 0.5, -0.5$ in different simulations, and 
the $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$.
The function $g$ describes how the environment $\epsilon_i(t)$ influences
$E_i(t)$. We used $g$ equal to $g_1$ or $g_2$ in different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise;
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that negatively impact populations,
and that occur only below a threshold of $\epsilon=0$; 
and $g_2$ represents effects that positively impact populations and 
that occur only above a threshold of $\epsilon=0$.
Negative values of $b$ correspond to overcompensating dynamics 
and positive values to undercompensating dynamics; larger $|b|$
means slower return to equilibrium after a disturbance.

For each $b$ and $g_i$, we simulated the model for $25000$ time
steps and retained the $E_i(t)$ for the final $2500$ time 
steps. We applied our model selection algorithms and 
nonparametric statistics (see section on methods for Q1) to these 
outputs to discover if the model could
produce non-normal copula structure and asymmetric tail dependence in the 
$E_i(t)$. Because $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have normal copula structure (they were drawn
from bivariate normal distributions), the Moran 
mechanism analyzed below 
does not operate here.
<!--***DAN: You may also want to have some additional commentary in the Discussion on
how the models and related models can be explored more comprehensively. In fact you may already have some of that, though you may want to beef it up. You may
want to insert a point at the end of the Methids section here again emphasizing
that these are simple models for an initial exploration, and we discuss in the 
Discussion how...-->

@@

To explore the importance of nonlinear dynamics for our proposed mechanism, 
we also considered a
nonlinear population model $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right)+\sigma\epsilon_i(t)\right)$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when it is strong, it brings the nonlinearities of the model into play. 
<!--***DAN: Didn't we use another value of r as well, and present results
on it? If so, probably should mention it here, in part so the exploration
seems a bit broader?-->

For each of the model setups above, for each of several copula families, 
<!--***DAN: I currently use C, SC, G, SG, J, SJ. Do I really need all those? 
Surely it would be enough to just do C and SC and then state that the others 
give similar results? And if I move one of the result figure from this section
into the main text, that will mean only one such fig in the sup mat, at least
for AR(1)? I mean, right now we have 18 figure in sup mat from this analysis,
and only part of one of them in main text! See if the results are actually similar across
S/SC, G/SG, J/SJ, and if so consider keeping only the S/SC results in either the
main text or the sup mat, and just stating the others were similar. Then you would presumably
move the full C fig into the main text, replacing the current partial version.-->
for each $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given Kendall
correlation $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were then computed for both noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted against $\tau$ for noises and populations. If the 
hypothesis from the Introduction was correct that characteristics of 
the copula structure of spatial dependence in an
ecological variable can be inherited from the copula structure of spatial 
dependence in an environmental variables, then 
plots should be similar for populations and noises.

<!--***DAN: How much of the material from Appendix S8 should be moved into
the main text? I am thinking a fair amount of it. Currrently there
is really no results on that topic at all in the main text, and that seems
a bit ridiculous and probably is why the referee said we treated this 
material in a "breezy" way.-->
The next mechanism we investigated is evolutionary, and pertains to bivariate 
trait data across species, e.g., our bird and mammal 
data. This mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
The hypothesis is that tail dependence occurs in evolutionary changes
in bivariate characters, and gives rise to tail dependence between the 
two character values across extant species.
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million 
independent draws from bivariate distributions showing one of five distinct types of copula 
structure: 1) extreme or 2) moderate left-tail dependence,
3) symmetric tail dependence, or 4) moderate or 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech}). 
All distributions had standard-normal marginals and Spearman 
correlation $0.875$ between components, so our simulations assess 
the impact of copula structure only. For 
each of the five copulas, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across phylogeny tips was assessed.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence in tip characters, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

Our model with asymmetric environmental effects 
produces outputs with visually apparent asymmetry 
of tail dependence, to an extent that depended on the value of $b$;
thus this is a mechanism that can produce non-normal copula structure and
asymmetric tail dependence across space in ecological variables.
<!--***DAN: move the following to the end of this part-->
This result is represented in Fig. \ref{MasterFigure} as the solid box around 
"Nonlinear environmental effects, Liebig's law" and the arrows labelled "A".
It is explained in the Discussion why the box and some of those arrows 
are solid, instead of dashed, although the results we 
present in this section are theoretical.
<!--***DAN: move the above to the end of this part-->
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ 
(Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetry of tail dependence was strong; for 
$b=\pm 0.5$, asymmetry was weaker but still apparent
(Figs \ref{SM-fig_ms_asym_b_0.5} - 
\ref{SM-fig_ms_asym_b_-0.5}).
<!--***DAN: combin all 3 sup mat figs on this subject into one sup mat fig-->
Lower-tail (respectively, upper-tail) spatial dependence in 
effects of noise ($g_1$, respectively, $g_2$) 
produced lower-tail (respectively, upper-tail) spatial 
dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) confirmed that
ecological outputs had non-normal copula structure, but did not always reflect 
asymmetry of tail dependence, perhaps because those methods measured
dependence in the extreme tails. Results using our nonparametric statistics, however,
reflected the visually apparent asymmetry (Tables \ref{SM-tab_stat_npa_asym}). 
<!--***DAN: Simplify as follows, maybe:
1) get rid of table S10 (you will have to change the methods so it no longer says we
do out model selection analysis here)
2) simplify table S11 so it only shows the cor results (also means a change to methods)
and then maybe bring the resulting table to main text?
-->

@@

The hypothesis that characteristics of 
spatial copula structure and tail dependence
in an ecological variable can be inherited from an environmental 
variable was largely confirmed; 
<!--***DAN: You are using language about confirming hypotheses, here and elsewhere,
but this may rub some people the wrong way because we have not confirmed anything
empirically, we are just doing models, and only a cursory exploration of each
model, at that! So change the language.-->
however, similarities between environmental- and 
ecological-variable copula structure were reduced when 
dynamics were strongly nonlinear.
Thus this is a mechanism that can produce non-normal copula structure and
asymmetric tail dependence 
<!--****DAN: Don't firget to replace @tail dependence@ by @tail asociation@ as appropriate-->
across space in ecological variables.
<!--***DAN: move the below to end of this part-->
This result is represented in Fig. \ref{MasterFigure} as the dashed box around 
"Moran effects" and the arrows labelled "B".
<!--***DAN: move the above to end of this part-->
For simulations using AR(1) models, copula 
statistics were always very similar for noise and populations 
(Fig \ref{fig_Cause4copula_scatter_C_kend_ar1}, 
\ref{SM-fig_Cause4copula_C_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1}),
though there were significant but minor differences for some statistics
and simulations.
For weak noise and our nonlinear model, copula statistics
were again very similar for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}).
Though there were again significant but small differences for some statistics
and simulations, the hypothesis was supported that copula 
structure in the noise translates
to similar but not identical copula structure in output populations. 
Since many ecological models are nonlinear, this result provides the 
expectation that there will be a Moran-effect-like correspondence 
between noise and model-output copula structure across space 
for typical ecological dynamics,
as long as noise is small enough that dynamics stay relatively close
to the model equilibrium. Results that hold for "weak noise" in this sense
are common in ecology. 
We repeated the analysis with $r=1.3$, with similar results. The deterministic 
one-habitat-patch Ricker model exhibits a monotonic approach to a stable equilibrium
at $K$ when $r<1$ (undercompensating dynamics, e.g., the value $r=0.5$ used initially), 
but exhibits an oscillatory approach to $K$ when $r>1$ (overcompensating
dynamics, e.g., $r=1.3$).
